\section{Unitary Coupled Cluster}

Within the traditional coupled-cluster framework, the ground electronic state is prepared by applying the CC operator to a reference state (usually Hartree-Fock).
\begin{equation*}
    \ket\psi = e^{\hat T} \ket{\phi_0}
\end{equation*}

Where $\hat T$ is the cluster excitation operator.

Quantum gates, however, must be unitary operators, so instead, we work within the UCC framework.

\begin{equation*}
    \ket\psi = e^{\hat T} \ket{\phi_0}
\end{equation*}

Where $\hat T$ is now an \textbf{anti-Hermitian} operator, and $e^{\hat T}$ is unitary.

In general, we can prepare exact electronic states by applying a sequence of $k$ parametrised unitary operators to our reference state.

\begin{equation*}
\begin{gathered}
    \ket\psi = \prod_i^k U_i(\theta_i) \ket{\phi_0} \\
    \text{Where $U_i(\theta_i)$ is a parametrised unitary operator}
\end{gathered}
\end{equation*}\smallskip

The parameters $\theta_i$ are then optimised to find the ground state energy.

General fermionic single and double excitation operators are defined as,
\begin{equation*}
    a_q^\dagger a_p \text{ and } a_r^\dagger a_s^\dagger a_q a_p
\end{equation*}

Exciting one electron from $p$ to $q$, and two electrons from $p, q$ to $r, s$ respectively.

Taking a linear combination of these, we obtain \textbf{anti-Hermitian} fermionic single and double excitation operators.
\begin{equation*}
\begin{gathered}
    \hat\kappa_p^q = a_q^\dagger a_p - a_p^\dagger a_q \\
    %
    \hat\kappa_{pq}^{rs} =
    a_r^\dagger a_s^\dagger a_q a_p - a_p^\dagger a_q^\dagger a_s a_r
\end{gathered}
\end{equation*}\smallskip

Such that upon exponentiating, we obtain \textbf{unitary} operators.

\begin{equation*}
    U^q_p = e^{\hat\kappa_p^q} \qquad
    %
    U_{pq}^{rs} = e^{\hat\kappa_{pq}^{rs}}
\end{equation*}

Recalling the Jordan-Wigner encoding for the creation and annhilation operators,

\begin{equation*}
    \hat a_j^+ = \frac{1}{2} (X - iY) \otimes Z^\rightarrow_{j-1} \qquad
    \hat a_j = \frac{1}{2} (X + iY) \otimes Z^\rightarrow_{j-1}
\end{equation*}

The anti-Hermitian fermionic single and double excitation operators $\kappa_p^q$ and $\kappa_{pq}^{rs}$
\begin{align*}
    F_p^q = \frac{i}{2} & (Y_p X_q - X_p Y_q) \prod_{k=p+1}^{q-1} Z_k \\
    %
    F_{pq}^{rs} = \frac{i}{8} (
      & X_p X_q Y_s X_r +
        Y_p X_q Y_s Y_r +
        X_p Y_q Y_s Y_r +
        X_p X_q X_s Y_r - \\
      & Y_p X_q X_s X_r -
        X_p Y_q X_s X_r -
        Y_p Y_q Y_s X_r -
        Y_p Y_q X_s Y_r )
    \prod_{k=p+1}^{q-1} Z_k
    \prod_{l=r+1}^{s-1} Z_l
\end{align*}

Multiplying by $\theta$ and exponentiating yields the parametrised unitary qubit operators,

\begin{equation*}
    U^q_p (\theta) =
    \text{exp} \left( i
    \frac{\theta}{2} (Y_p X_q - X_p Y_q) \prod_{k=p+1}^{q-1} Z_k \right)
\end{equation*}

\begin{equation*}
    U^{rs}_{pq} (\theta) = \text{exp} \left( i \frac{\theta}{8} (
    X_p X_q Y_s X_r
    + \dots -
    Y_p Y_q Y_s X_r -
    Y_p Y_q X_s Y_r )
    \prod_{k=p+1}^{q-1} Z_k
    \prod_{l=r+1}^{s-1} Z_l
    \right)
\end{equation*}

To summarise, we constructed anti-Hermitian single and double excitation operators from a linear combination of fermionic excitation operators,

\begin{equation*}
    \hat\kappa_p^q = a_q^\dagger a_p - a_p^\dagger a_q \qquad
    %
    \hat\kappa_{pq}^{rs} =
    a_r^\dagger a_s^\dagger a_q a_p - a_p^\dagger a_q^\dagger a_s a_r
\end{equation*}\smallskip

We then mapped these to qubit operators using the Jordan-Wigner transformation,
\begin{equation*}
    \hat\kappa_p^q \xrightarrow{\text{JW}} F_p^q \qquad\qquad
    \hat\kappa_{pq}^{rs} \xrightarrow{\text{JW}} F_{pq}^{rs}
\end{equation*}

And finally, we exponentiated to yield the parametrised unitary qubit operators.
\begin{equation*}
    U^q_p (\theta) = e^{\theta^q_p F_p^q} \qquad
    %
    U^{rs}_{pq}(\theta) = e^{\theta_{pq}^{rs} F_{pq}^{rs}}
\end{equation*}

