\chapter{Conclusion}%
\label{conclusion}

\section{Summary}

% The research presented in this thesis focuses on the diagrammatic representation of fermionic excitation operators using the ZX calculus. Working within the DISCO-VQE framework, our initial goal was to identify a generalised structure for these excitation operators, anticipating that this might enable us to gain deeper insights into the nature of the correlations present in molecular systems. We also aimed to rationalise the expectation energy-equivalent ans채tze produced by the DISCO-VQE algorithm, potentially optimising the algorithm by minimising the redundancies in its outputs. Furthermore, we sought to optimise the implementation of the UPS ansatz itself by reducing circuit depth.

% Through our exploration, we demonstrated the expressive power of the ZX calculus in representing and manipulating fermionic excitation operators. By using the diagrammatic rewrite rules afforded by the ZX calculus in Chapter \ref{zx-calculus} and the commutation relations developed in Section \ref{pauli-gadgets}, we were able to identify previously unknown non-trivially commuting excitation operators used in the DISCO-VQE algorithm. We used these commutation relations to replicate the work done by Yordanov \textit{et al} in the ZX calculus framework. In this way, we were able to identify a generalised representation for fermionic excitation operators as controlled rotations. In Chapter \ref{controlled-rotations}, we developed a representation for higher order controlled rotations used in the ZX calculus.

% Our findings showed that by identifying commuting pairs of fermionic excitation operators, we can effectively reduce the number of unique ans채tze required to explore the Hilbert space. This reduction not only optimises the algorithm but also enhances our understanding of the underlying quantum states and their correlations. Additionally, the insights gained from the ZX calculus allowed us to propose modifications to the DISCO-VQE algorithm, further improving its efficiency and effectiveness.

% The application of the ZX calculus has proven to be a powerful tool in this endeavour, offering both theoretical insights and practical improvements. Future research can build on these findings to develop more advanced algorithms and techniques, further pushing the boundaries of what is possible in quantum simulation of molecular systems.

%%%

Our research focuses on the diagrammatic representation of fermionic excitation operators using the ZX calculus within the DISCO-VQE framework. Our primary goals were to gain deeper insights into the correlation present in molecular systems, rationalise the redundancies among expectation energy-equivalent ans채tze, and optimise the UPS ansatz by reducing circuit depth.

The first four chapters of this thesis lay the groundwork for the findings presented in Chapter \ref{excitation-operators}. In Chapter \ref{background}, we began by introducing Electronic Structure Theory, quantum computing, and VQE algorithms. We discussed DISCO-VQE as the specific algorithm that forms the framework for the research conducted in this thesis, we introduced the UPS ansatz, explaining how it can be used to parametrically explore the Hilbert space of possible states to find approximate solutions for the ground state energy. In Chapter \ref{zx-calculus}, we introduced the ZX calculus framework, its basic generators, and its rewrite rules. Then, in Chapter \ref{pauli-gadgets}, we introduced Pauli gadgets and their representation in the ZX calculus. We explain that they can be used as the building blocks for the fermionic excitation operators used to simulate fermionic systems. In this context, we outlined the set of rules describing how Pauli gadgets interact with each other in Section \ref{commutation-relations}. Then, in Section \ref{clifford-commutation-relations}, we developed a diagrammatic method to reproducibly derive the interactions of Pauli gadgets with any Pauli or Clifford gate. In Chapter \ref{controlled-rotations}, we extended the work of Yeung \cite{Yeung2020} on singly-controlled rotations to develop a representation of higher-order controlled rotations in the ZX calculus. We later used this result in our attempt to identify a generalised structure for fermionic excitation operators in the ZX calculus.

In Chapter \ref{excitation-operators}, we build on this foundational work and address our primary goals by first outlining the representation of fermionic excitation operators in terms of Pauli gadgets in Section \ref{implementing-excitation-operators} and explaining how they can be reliably implemented as quantum circuits with optimal circuit depth. We demonstrated the advantages of representing these operators in this architecture-independent manner, illustrating how the gadget commutation rules derived in Section \ref{commutation-relations} can be used to provide insights into the structure of the UPS ansatz. In this context, we discussed the expectation energy-equivalent UPS ans채tze generated by the DISCO-VQE algorithm in Section \ref{operator-commutations} and proposed a rationalisation for the redundancy in their representation. We identified several previously unknown non-trivially commuting excitation operators in the operator pool and proposed a modification to the DISCO-VQE algorithm to account for these redundancies. Finally, we demonstrated the generality of this result, explaining how these commutation relations hold for any molecular system with a given number of spin orbitals. In Section \ref{operator-controlled-rotations}, we used the commutation relations developed in Section \ref{clifford-commutation-relations} to replicate the results by Yordanov \textit{et al} and Kornell \textit{et al} using the ZX calculus. By conjugating the excitation operators with a Clifford subcircuit, we obtained a controlled rotation, whose representation in the ZX calculus we derived in Chapter \ref{controlled-rotations}. These derivations validate the expressive power of studying excitation operators diagrammatically. We argue that by expressing the excitation operators in terms of controlled rotations, we uncover the underlying nature of these operators, providing an alternate perspective on their behaviour.

In Chapter \ref{zxfermion}, we introduce the software package ZxFermion that we built to explore research ideas related to circuits of Pauli gadgets. By encoding the commutation relations developed previous chapters, we were able to straightforwardly replicate the results of Yordanov \textit{et al} and Kornell \textit{et al}, demonstrating a noteworthy acceleration in research pace. By providing an accessible tool for studying the interactions of Pauli gadgets, we hope to facilitate the research done by future chemists wanting to get started in quantum computing, hence lowering the barrier to entry.

\section{Future Work}

